---
output: github_document
---

# surveysd

[![Build Status](https://travis-ci.org/statistikat/surveysd.svg?branch=master)](https://travis-ci.org/statistikat/surveysd)
<!--[![Coverage Status](https://coveralls.io/repos/github/statistikat/surveysd/badge.svg?branch=master)](https://coveralls.io/github/statistikat/surveysd?branch=master)-->
<!--[![CRAN](http://www.r-pkg.org/badges/version/surveysd)](https://CRAN.R-project.org/package=surveysd)-->
<!--[![Downloads](http://cranlogs.r-pkg.org/badges/surveysd)](https://CRAN.R-project.org/package=surveysd)-->
<!--[![Mentioned in Awesome Official Statistics ](https://awesome.re/mentioned-badge.svg)](http://www.awesomeofficialstatistics.org)-->

This is the development place for the R-package `surveysd`. This package can be used to estimate 
the standard deviation of estimates in complex surveys using bootstrap weights.

## Installation

This package can be installed like any other R package on github via `install_github`

```{r, eval = FALSE}
devtools::install_github("statistikat/surveysd")
```

## Syntax example

This example showcases the three most important functions in the surveysd package: `draw.boostrap`,
`recalib` and `calc.stError`.

### Load dummy data

```{r}
library(surveysd)

set.seed(1234)
eusilc <- demo.eusilc(n = 2, prettyNames = TRUE)
``` 

### Draw bootstrap replicates

Use stratified resampling without replacement to generate 10 samples. Those samples are
consistent with respect to the reference periods.

```{r}
dat_boot <- draw.bootstrap(eusilc, REP = 10, hid = "hid", weights = "pWeight", 
                           strata = "region", period = "year")
```
   
### Calibrate bootstrap replicates

Calibrate each sample according to the distribution of `gender` (on a personal level) and `region` 
(on a household level). 

```{r}
dat_boot_calib <- recalib(dat_boot, conP.var = "gender", conH.var = "region")
```

### Estimate with respect to a grouping variable

Estimate relative amount of persons at risk of poverty per period and `gender`.

```{r}
err.est <- calc.stError(dat_boot_calib, var = "povertyRisk", fun = weightedRatio,
                        group = "gender", period.diff = NULL, period.mean = NULL)
err.est$Estimates
```

The output contains estimates (`val_povertyRisk`) as well as standard errors (`stE_povertyRisk`)
measured in percent.

### Estimate with respect to several variables

Estimate relative amount of persons at risk of poverty per period for each `region`, 
`gender`, and combination of both.

```{r}
group <- list("gender", "region", c("gender", "region"))
err.est <- calc.stError(dat_boot_calib, var = "povertyRisk", fun = weightedRatio,
                        group = group, period.diff = NULL, period.mean = NULL)
head(err.est$Estimates)
## skipping 54 more rows
```
